---
title: "Functions making up a Force-Field"
author: 
- Mauricio Esguerra PhD ^[Department of Cell and Molecular Biology,  Uppsala University]
#date: "March 12, 2018"
header-includes:
  - \usepackage[T1]{fontenc}
  - \usepackage[scaled]{helvet}
  - \renewcommand*\familydefault{\sfdefault}
output:
  pdf_document:
    toc: true
    number_sections: true
  html_document:
    df_print: paged
  word_document: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Bonds as Harmonic Potential Functions

```{r, echo=TRUE}
################################################################################
# BONDS AS HARMONIC POTENTIALS
################################################################################
#Force constants for carbon single, double, and triple bonds
k1=222.50; k2=300; k3=450

#Equilibrium distances for carbon single, double, and triple bonds in Ångström
b1=1.5300; b2=1.3; b3=1.0

#Sampled variable
#b=runif(100, 0.5, 2.5)
b=seq(0.0, 2.5, ((2.5-0.0)/50))

#Harmonic Potential function
vbond1 <- k1*(b-b1)^2
vbond2 <- k2*(b-b2)^2
vbond3 <- k3*(b-b3)^2

xlim=range(0.0, 2.5) 
ylim=range(0,400)
#plot(b, vbond1, pch=23, type="o", col="blue", cex=0.8,
#     xlab = " ", main="", xlim = xlim, ylim = ylim, ylab= " ",
#     panel.first = grid(), axes=FALSE, xaxs="i", yaxs="i")
plot(b, vbond1, pch=23, type="o", col="blue", cex=0.6,
     ylab="potential (bond distance)",main="", xlim = xlim, ylim = ylim,
     panel.first = grid(), xaxs="i", yaxs="i")
lines(b, vbond2, col="red",type="o", cex=0.6)
lines(b, vbond3, col="gray",type="o", cex=0.6)
#axis(1, cex.axis=1.0, at=seq(0, 2.5,0.5))
#axis(2, cex.axis=1.0, at=seq(0, 400,50),las=1) 
#box(which="plot",col="black")
legend("top",col=c("blue","red","gray"),lty=1,
       legend=c("single C bond","double C bond","triple C bond"), cex=0.8)
```

# Bonds as Morse Potentials

```{r, echo=TRUE}
################################################################################
# BONDS AS MORSE (ANHARMONIC) POTENTIALS
################################################################################
#Force constants for carbon single, double, and triple bonds
k1=222.50; k2=300; k3=450

#Equilibrium distances for carbon single, double, and triple bonds in Ångström
b1=1.5300; b2=1.3; b3=1.0

#Well depths (De) for anharmonic potential
De1 <- 10000
De2 <- 10000
De3 <- 10000
#
alpha1 <- sqrt(k1/(2*De1))
alpha2 <- sqrt(k2/(2*De2))
alpha3 <- sqrt(k3/(2*De3))
#Sampled variable
#b=runif(100, 0.5, 2.5)
b=seq(0.0, 2.5, ((2.5-0.0)/50))

#Harmonic Potential function
vbond1 <- De1*(1-exp(-alpha1*(b-b1)))^2
vbond2 <- De2*(1-exp(-alpha2*(b-b2)))^2
vbond3 <- De3*(1-exp(-alpha3*(b-b3)))^2

xlim=range(0.0, 2.5) 
ylim=range(0,400)
#plot(b, vbond1, pch=23, type="o", col="blue", cex=0.8,
#     xlab = " ", main="", xlim = xlim, ylim = ylim, ylab= " ",
#     panel.first = grid(), axes=FALSE, xaxs="i", yaxs="i")
plot(b, vbond1, pch=23, type="o", col="blue", cex=0.6,
     ylab="potential (bond distance)",main="", xlim = xlim, ylim = ylim,
     panel.first = grid(), xaxs="i", yaxs="i")
lines(b, vbond2, col="red",type="o", cex=0.6)
lines(b, vbond3, col="gray",type="o", cex=0.6)
#axis(1, cex.axis=1.0, at=seq(0, 2.5,0.5))
#axis(2, cex.axis=1.0, at=seq(0, 400,50),las=1) 
#box(which="plot",col="black")
legend("top",col=c("blue","red","gray"),lty=1,
       legend=c("single C bond","double C bond","triple C bond"), cex=0.8)
```



# Dihedral Angle Potentials as Cosines or a Sum of Cosines

```{r, echo=TRUE}
################################################################################
# DIHEDRAL ANGLES AS COSINES OR FOURIER SERIES
################################################################################

#k3=0.1525; k2=0.3; k1=0.6 # k=Vo/2, so the barrier for rotation is Vo.
k3=0.1525; k2=k3; k1=k3
n3=3; n2=2; n1=1    # number of barriers or minima.
delta=0
phi=seq(0, 2*pi, ((2*pi)/100))

#angles are given in radians so have to convert on ploting to degrees.
vdihe1 <- k1*(1+(cos(n1*phi-delta)))
vdihe2 <- k2*(1+(cos(n2*phi-delta)))
vdihe3 <- k3*(1+(cos(n3*phi-delta)))

vdihet <- vdihe1+vdihe2+vdihe3

xlim=range(0, 360) 
ylim=range(0,1.0)
plot(phi*(360/(2*pi)), vdihe1, pch=23, type="o", col="blue", cex=0.5,
     xlab = "dihedral angle alkane", main="", xlim = xlim, 
     ylim = ylim, ylab= "dihedral potential",
     panel.first = grid(), axes=FALSE, xaxs="i", yaxs="i")
lines(phi*(360/(2*pi)), vdihe2, col="red",type="o", cex=0.5)
lines(phi*(360/(2*pi)), vdihe3, col="gray",type="o", cex=0.5)
lines(phi*(360/(2*pi)), vdihet, col="green",type="o", cex=0.5)
axis(1, cex.axis=1.0, at=seq(0, 360,30))
axis(2, cex.axis=1.0, at=seq(0, 2,0.2),las=1) 
box(which="plot",col="black")
legend("top",col=c("blue","red","gray","green"),lty=1,
       legend=c("nmin=1","nmin=2","nmin=3","sum"), cex=0.8)
```


# NON-BONDED (INTERMOLECULAR) INTERACTIONS  

## Lennard-Jones potential  
```{r, echo=TRUE}
################################################################################
# NON-BONDED (INTERMOLECULAR) INTERACTIONS
################################################################################
# Lennard-Jones potential
epsiloni=-0.0780 #kcal/mole
epsilonj=-0.0220 #kcal/mole
epsilonij=sqrt(epsiloni*epsilonj) #Lorentz-Berthelot Lorentz=geometric
rmini=2.040*2   #Ångström
rminj=1.32*2    #Ångström
rminij=((rmini+rminj)/2)          #Lorentz-Berthelot Berthelot=aritmetic
rij=seq(0.0, 7.0, ((7.0-0.0)/100))
Aij= ((rminij/rij)**12)
Bij= (2*(rminij/rij)**6)
lj=epsilonij*( Aij - Bij)

xlim=range(2.5, 6.0) 
ylim=range(-0.05,0.05)
plot(rij,epsilonij*Aij, pch=23, type="o", col="blue", cex=0.5,
     xlab = expression("r"~(ring(A))), main="", xlim = xlim, ylim = ylim, ylab= "U(r)",
     panel.first = grid(), axes=FALSE, xaxs="i", yaxs="i")
lines(rij,-epsilonij*Bij, pch=23, type="o", col="red", cex=0.5,
      xlab = " ", main="", xlim = xlim, ylim = ylim, ylab= " ",
      panel.first = grid(),  xaxs="i", yaxs="i")
lines(rij,lj, pch=23, type="o", col="gray", cex=0.5,
      xlab = " ", main="", xlim = xlim, ylim = ylim, ylab= " ",
      panel.first = grid(), xaxs="i", yaxs="i")
abline(v=rminij,col="black",lty=2, lwd=2)
axis(1, cex.axis=1.0, at=seq(2.5, 6.0, 0.5))
axis(2, cex.axis=1.0, at=seq(-0.05, 0.05, 0.01),las=1) 
legend("topright",col=c("blue","red","gray"),lty=1,
       legend=c(expression("repulsive"~(r[0]/r)^12),
                expression("attractive"~-(r[0]/r)^6),"L-J"), cex=0.8)
text(c(3.1), c(-0.045), labels=c("-0.0412"), cex= 0.8)
box(which="plot",col="black")
```

## Electrostatic (Coulomb) potential
It is Coulomb's law in AKMA units, ie length in Å, energy in kcal/mol and 
charge in |e|.
A relative dielectric permittivity (dimensionless) is allowed for in the 
denominator.
```{r, echo=TRUE}
# electrostatic potential
# It is Coulomb's law in AKMA units, ie length in Å, energy in kcal/mol and charge in |e|.
# A relative dielectric permittivity (dimensionless) is allowed for in the denominator 
# (keyword EPS, default value 1.0). Numerically this boils down to:
# E= 332*(qi*qj)/(rij*eps)
# Kcoul = 1/(4*pi*epsilon-zero)
# epsilon-zero = 8.8542 x 10^-12 C^2 m^-1 J^-1
# 1 m = 10^10 A
# 1 C = 1.6022 x 10 ^19 esu
# 1 J = 6.0221 x 10 23 / 4184 kcal mol^-1
# Kcoul = 1 J m / 4*pi*(8.8542 x 1o ^-12)C^2 == 332 kcal * mol^-1 * esu^-2
qi=-0.27  # carbon charge
qj=0.09   # hydrogen charge
e=1 # dielectric constant 1 for vacuum, 80 for water, 4 for protein interior 
rij=seq(0.0, 7.0, ((7.0-0.0)/100))
Kcoul=332
velecc= Kcoul*(qi*qi)/(e*rij)
velech= Kcoul*(qi*qj)/(e*rij)


xlim=range(0, 6.0) 
ylim=range(-50,50)
plot(rij,velecc, pch=23, type="o", col="blue", cex=0.5,
     xlab = expression("r"~(ring(A))), main="", xlim = xlim, ylim = ylim, 
     ylab= "U(r)",
     panel.first = grid(), axes=FALSE, xaxs="i", yaxs="i")
lines(rij,velech, pch=23, type="o", col="red", cex=0.5,
      xlab = " ", main="", xlim = xlim, ylim = ylim, ylab= " ",
      panel.first = grid(),  xaxs="i", yaxs="i")
#abline(v=rminij,col="gray",lty=2, lwd=2)
axis(1, cex.axis=1.0, at=seq(0.0, 6.0, 0.5))
axis(2, cex.axis=1.0, at=seq(-50, 50, 10),las=1) 
legend("topright",col=c("blue","red"),lty=1,legend=c("C(-0.27) .. C(-0.27)",
      "C(-0.27) .. H(+0.09)"), cex=0.8)
box(which="plot",col="black")
```


